```{r}
source("./all_artifacts.r")
```

### Taxonomic classifications
```{r}
blast <- lapply(
    get_artifact_data("./results/3-Classified", id_key, "BLAST_All"),
    parse_taxonomy
)
sklearn <- lapply(
    get_artifact_data("./results/3-Classified", id_key, "Sklearn"),
    parse_taxonomy
)
sk_merged <- read_qza("./results/3-Classified/Merged-Sklearn.qza")$data %>%
    parse_taxonomy()
```
Simply by counting the number of NA entries in the taxonomy columns, the sklearn
classifier consistently outperforms the traditional blast approach. Furthermore, the sklearn is faster and consumes less memory (aside from the initial training step, which I could not do)

The `known_taxon` function collapses the taxonomy table into the first identification. A similar function will be used to collapse it into at least genus identifications for use with farprotax.

```{r}
# Cleaning data for logistic regression
all <- read_qza("./results/2-OTUs/Merged-otuFreqs.qza")$data
sk_merged <- read_qza("./results/3-Classified/Merged-Sklearn.qza")$data %>%
    parse_taxonomy()

orders <- merge_with_id(all, sk_merged, level = 2) %>%
    filter(!(is.na(taxon))) %>%
    group_by(taxon) %>%
    summarise(across(everything(), sum))

to_log <- orders %>%
    rel_abund(., taxon) %>%
    as.data.frame() %>%
    t() %>%
    as.data.frame() %>%
    rename_with(~ as.character(as.vector(orders$taxon)), all_of(colnames(.))) %>%
    mutate(ids = rownames(.)) %>%
    filter(!(ids == "taxon"))  %>%
    inner_join(metadata, ., by = join_by(x$sample.id == y$ids)) %>%
    as_tibble()
write.csv(to_log, "logistic_reg.csv", row.names = FALSE)

# Lots of false positive hits?
not_bacteria <- c(
    "Arthropoda", "Nanoarchaeota", "Diatomea", "Altioarchaeota",
    "Ascomycota", "Basidiomycota", "Cercozoa", "Ciliophora"
)
summed <- sapply(names(id_key), function(x) {
    rowSums(orders[, grep(x, colnames(orders)), drop = FALSE])
}) %>%
    as_tibble() %>%
    mutate(taxon = orders$taxon) %>%
    relocate(taxon) %>%
    filter(!(grepl("[0-9]", taxon))) %>%
    rel_abund(., taxon) %>%
    pivot_longer(., -taxon)

stacked <- summed %>% ggplot(., aes(x = name, y = value, fill = taxon)) +
    geom_bar(stat = "identity") +
    scale_color_paletteer_d("pals::glasbey")
```

### Phylogenetic trees
```{r}
fasttree <- get_artifact_data(
    "./results/6-RootedTrees", id_key,
    "FastTree_RootedTree"
)
iqtree <- get_artifact_data(
    "./results/6-RootedTrees", id_key,
    "IQTREE_RootedTree"
)
```



### Import frequency tables
The frequency tables are stored in BIOM format: OTUs x sites
```{r}
otu_freqs <- lapply(
    get_artifact_data("./results/2-OTUs", id_key, "otuFreqs"),
    as.data.frame
)
```

The following snippet maps otu ids the genus-level identifications for use with the farprotax database, writing csv files for each.
```r
otu_genus <- list()
for (id in names(id_key)) {
    otu_genus[[id]] <- to_genus_csv(otu_freqs[[id]], blast[[id]])
}
genus_combined <- combine_freqs(otu_genus, taxon)
# write.csv(combined, "genus_otu_tables.csv", row.names = FALSE)
```

## Diversity analyses

### Alpha diversity
```{r}
alpha <- get_artifact_data("./results/7-Diversity",
    id_key,
    extension = "",
    metric_list = alpha_metrics
)
pi_evenness <- data.frame(row.names = 1:10)
for (site in names(id_key)) {
    pi_evenness[[id_key[[site]]]] <- alpha[[site]]$pi[, 1]
}
even_plot <- pi_evenness %>%
    gather() %>%
    ggplot(aes(x = key, y = value)) +
    geom_boxplot(fill = "lightcoral") +
    labs(x = "Site", y = "Evenness", title = "Pielou evenness")
ggsave("./figures/Pielou_evenness.png", even_plot, width = 20, height = 10)
```
From the box plot, we can see a few of the sites having similar evenness, particularly the Barrow mountain sites, Bihor mountains and Catriona snow. Whether or not the difference in Pielou evenness is statistically significant can be tested with Kruskal wallis
```{r}
pi_evenness %>%
    select(c(
        "Barrow mountain high", "Barrow mountain low",
        "Bihor mountains", "Catriona snow"
    )) %>%
    kruskal.test()
```
The p-value is 0.2744, so the difference in evenness is not statistically significant


### Beta diversity
All ordination plots will be carried out with PCOA
```{r}
beta <- get_artifact_data("./results/7-Diversity",
    list(Merged = NULL),
    extension = "",
    metric_list = beta_metrics
)
pcoa2D <- get_artifact_data("./results/8-Analysis",
    list(Merged = NULL),
    extension = "PCOA-2D_",
    metric_list = beta_metrics
)
pcoa2D_merged <- lapply(pcoa2D$Merged, metadata_merge_pcoa, metadata = metadata)
pcoaja <- plot_pcoa(pcoa2D_merged$ja, "Location",
    title = "Jaccard distance"
)
pcoabc <- plot_pcoa(pcoa2D_merged$bc, "Location",
    title = "Bray Curtis"
)
pcoauu <- plot_pcoa(pcoa2D_merged$uu, "Location",
    title = "Unweighted Unifrac",
)
pcoawn <- plot_pcoa(pcoa2D_merged$wn, "Location",
    title = "Weighted Normalized Unifrac ",
)
pcoa_arrange <- ggarrange(pcoabc, pcoauu, pcoawn,
    ncol = 3,
    common.legend = TRUE,
    legend = "right"
)
ggsave("./figures/pcoas.png",
    plot = pcoa_arrange,
    dpi = "retina", width = 20, height = 10
)
```
The pcoa plots depict clearly how the choice of distance metric affects the clustering of samples.
- Clustering in the Bray Curtis plot represents sites sharing many of the same species and in similar abundances. The big cluster falls apart once we factor the evolutionary distance between OTUs, shown by the Unifrac metrics.
    - This implies that although some taxa are shared, the unique taxa are a evolutionary distant (essentially have very different DNA) from the shared ones.
- Once we add weightings by abundance though, new clusters form, indicating there are many more common taxa within the groups than there are unique taxa.
- Even partitioning sites by habitat type - glacier and permafrost, doesn't help

Admittedly, there is no clear consensus on what metric to choose when the results disagree strongly with each other. For the scope of this project, I will focusing the rest of the analyses only on the cluster depicted in the unweighted unifrac plot consisting of Cryoconite, Bihor Mountain caves, Yakutia, East Iceland glaciers, Catriona snow, and Greenland ice.
```{r}
# Filter the sites
keep <- c("BhM", "CrC", "CaS", "CeY", "GrI", "EaI")
keep2 <- c("ViS", "BrL", "BrH", "SvG")
keep3 <- c("CaS", "CeY", "BhM", "EaI")
filter_uu <- filter_dm(beta$Merged$uu, keep)
filtered_meta <- filter_meta(metadata, keep)
filter_wn <- filter_dm(beta$Merged$wn, keep2)
filtered_meta2 <- filter_meta(metadata, keep2)
filter_bc <- filter_dm(beta$Merged$bc, keep3)
filtered_meta3 <- filter_meta(metadata, keep3)
```

#### PERMANOVA & ANOSIM
Within the chosen cluster, its visually unclear whether or not the differences between sites are statistically signif
```{r}
adonis2(filter_uu ~ filtered_meta$Location)
# adonis2(filter_wn ~ filtered_meta2$Location)
# adonis2(filter_bc ~ filtered_meta3$Location)
anosim(filter_uu, grouping = filtered_meta$Location)
```

### Functional analysis
The raw output tables here are the absolute abundances of molecular functions in each site using the KEGG Orthology database (KO). PICRUSt2 integrates with other databases, such as Enzyme Commission numbers (EC), but I've chosen to use the KO database because the PICRUSt2 authors describe how they've improved on the database integration compared to PICRUSt1.
As PICRUSt2 failed to analyze the merged dataset, I will combine the output tables from each sample.
```{r}
ko_all <- ko %>%
    # Merge the PICRUSt2 tables
    reduce(merge, by = "function", all = TRUE) %>%
    as_tibble() %>%
    replace(is.na(.), 0) %>%
    rel_abund(., `function`) %>% # Convert to relative abundances
    as_tibble() %>%
    sites_x_func() # Transpose into sites x function format

# Compute bray curtis distance, then plot pcoa
bc_func <- vegdist(ko_all, method = "bray")
pcoa_bc_func <- bc_func %>%
    wcmdscale(k = 2) %>%
    metadata_merge_pcoa(metadata, ., functions = TRUE)
# Compute jaccard distance
ja_func <- vegdist(ko_all, method = "jaccard")
pcoa_ja_func <- ja_func %>%
    wcmdscale(k = 2) %>%
    metadata_merge_pcoa(metadata, ., functions = TRUE)

# Plot and compare with ordination on taxonomy
plot_ja_func <- plot_pcoa(pcoa_ja_func, "Location",
    title = "Jaccard distance",
    subtitle = "From biological functions", functions = TRUE
)
plot_bc_func <- plot_pcoa(pcoa_bc_func, "Location",
    title = "Bray Curtis",
    subtitle = "From biological functions", functions = TRUE
)
func_compare <- ggarrange(plot_ja_func, pcoaja, plot_bc_func, pcoabc,
    ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom"
)
ggsave("./figures/func_vs_tax.png",
    plot = func_compare,
    dpi = "retina", width = 20, height = 10
)
```

### Differential abundance analysis
TODO: finish Now we will perform tests on the abundance of identified OTUs within the previously selected sites. I will be using the Analysis of Microbiomes with Bias correction test (ANCOM-BC), 
The test can be thought of as ANOVA in that it identifies taxa that are differentially abundant between at least two groups across a set
- ANCOM-BC returns five metrics for each OTU:
    - Log fold change across different sites
    - W-score: log fold change across different sites divided by standard error. This is ANCOM-BC's test statistic
    - p value: whether or not the log fold change for a given OTU is statitically significant
    - q value: the adjusted p-values
    - Standard error
```r
# Write the filtered metadata to a file for filtering
write_delim(filtered_meta, "to_keep.tsv", delim = "\t")
```
```bash
# One of the assumptions that ANCOM-BC makes is that fewer than 25% of OTUs are changing between groups, so as with ANOSIM and PERMANOVA, we keep only the sites that appear close together through beta diversity
qiime feature-table filter-samples \
--i-table 2-OTUs/Merged-otuFreqs.qza \
--m-metadata-file ../to_keep.tsv \
--o-filtered-table 8-ANCOM-BC/Merged_filtered-otuFreqs.qza

# Collapse into the taxonomy into order-level identifications
qiime taxa collapse \
--i-table 8-ANCOM-BC/Merged_filtered-otuFreqs.qza \
--i-taxonomy 3-Classified/Merged-Sklearn.qza \
--p-level 4 \
--o-collapsed-table 8-ANCOM-BC/Merged_collapsed-otuFreqs.qza

qiime composition ancombc \
--i-table 8-ANCOM-BC/Merged_collapsed-otuFreqs.qza \
--m-metadata-file ../ds_metadata.tsv \
--p-formula Location \
--o-differentials 8-ANCOM-BC/Merged-ancombc.qza

qiime tools export \
--input-path 8-ANCOM-BC/Merged-ancombc.qza \
--outpath-path 8-ANCOM-BC/export
```

```{r}
# Import all ancombc results
wrong_tax <- c("Unassigned", "Arthropoda", "Bacteria", "Insecta")
abc_path <- "./results/8-ANCOM-BC"
# Filter by statistically significant results only
ancombc_q_val <- import_ancom("q_val_slice.csv", abc_path)
ancombc_se <- import_ancom("se_slice.csv", abc_path)
all_sig <- (ancombc_q_val[, -1] < 0.05) %>%
    apply(., 1, all)
filtered_q <- ancombc_q_val %>% filter(all_sig)
# Obtain log-fold change divided by standard error
ancombc_lfc <- import_ancom("lfc_slice.csv", abc_path) %>% filter(all_sig)
ancombc_lfc$id <- ancombc_lfc$id %>%
    lapply(., collapse_tax, level = 4) %>%
    unlist()
ancombc_lfc <- ancombc_lfc %>%
    group_by(id) %>%
    summarise(across(everything(), sum)) %>%
    filter(!(id %in% wrong_tax))
abc_lfc <- ancombc_lfc %>%
    pivot_longer(., -id) %>%
    ggplot(aes(x = id, y = value, fill = name)) +
    geom_bar(stat = "identity", position = position_dodge())
# ancombc_q_val <- import_ancom("w_slice.csv", abc_path)
# all_sig2 <- (ancombc_q_val[, -1] < 0.05) %>%
#     apply(., 1, all)
# ancombc_q_val <- ancombc_q_val %>% filter(all_sig2)
# ancombc_q_val$id <- ancombc_q_val$id %>%
#     lapply(., collapse_tax, level = 5) %>%
#     unlist()
```

```{r}
# Manual ancombc
tax_info <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
filtered_otus <- read_qza("./results/2-OTUs/Merged-otuFreqs.qza")$data %>%
    as.data.frame() %>%
    select(matches(keep))
m <- merge(sk_merged, filtered_otus, by = 0)
# TODO: Drop all rows with 0s
filtered_otus <- m %>% select(!(matches(tax_info)))
tax <- m %>% select(matches(c("Row.names", tax_info)))
# tse <- TreeSummarizedExperiment(assays = list(otus = filtered_otus))
```

