```{r}
source("./all_artifacts.r")
```
### Phylogenetic trees
```{r}
fasttree <- get_artifact_data(
    "./results/6-RootedTrees", id_key,
    "FastTree_RootedTree"
)
iqtree <- get_artifact_data(
    "./results/6-RootedTrees", id_key,
    "IQTREE_RootedTree"
)

# Sample tree for Villum station site
# Still need to improve separation of tip labels
villum <- replace_tips(iqtree$ViS, blast$ViS)
plot(villum)
```

### Taxonomic classifications
```{r}
blast <- lapply(
    get_artifact_data("./results/3-Classified", id_key, "BLAST_All"),
    parse_taxonomy
)
```
The `known_taxon` function collapses the taxonomy table into the first identification. A similar function will be used to collapse it into at least genus identifications for use with farprotax.

### Import frequency tables
The frequency tables are stored in BIOM format: OTUs x sites
```{r}
otu_freqs <- lapply(
    get_artifact_data("./results/2-OTUs", id_key, "otuFreqs"),
    as.data.frame
)
```

The following snippet maps otu ids the genus-level identifications for use with the farprotax database, writing csv files for each.
```{r}
otu_genus <- list()
for (id in names(id_key)) {
    otu_genus[[id]] <- to_genus_csv(otu_freqs[[id]], blast[[id]])
}
combined <- bind_rows(otu_genus) %>%
    arrange(taxon) %>%
    group_by(taxon) %>%
    summarise(across(everything(), sum))
combined[is.na(combined)] <- 0.0
write.csv(combined, "genus_otu_tables.csv", row.names = FALSE)
```

## Importing PCOA artifacts

```{r}
```
